Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
added references to intro

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al. 2019, **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (**REFS(KAT)**)--for the observed greater drought growth reductions of large trees. First, taller trees facea greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (), and this may become a greater liability during drought (**REFS(KAT)**). Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (**REFS(KAT)**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS(KAT)**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al. 2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; **REF**) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs (**REFS**), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). Their ability to predict tree performance under drought remains untested.

>>>>

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al. 2019, **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (*Pfeifer, E. M., Hicke, J. A. & Meddens, A. J. H. Observations and modeling of aboveground tree carbon stocks and fluxes following a bark beetle outbreak in the western United States. Glob. Change Biol. 17, 339–350 (2011).*)--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (* McDowell, N. G., Bond, B. J., Hill, L., Ryan, M. G. & Whitehead, D. in Size and age related changes in tree structure and function (eds Meinzer, F.C. & Niinemets, U.) 255–286 (Springer Publishing, 2011); McDowell, N. G. & Allen, C. D. Darcy’s law predicts widespread forest mortality under climate warming. Nat. Clim. Change 5, 669–672 (2015); Ryan, M. G., Phillips, N. & Bond, B. J. The hydraulic limitation hypothesis revisited. Plant Cell Environ. 29, 367–381 (2006).*), and this may become a greater liability during drought ([Zhang et al. 2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)). Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@suarez_factors_2004]; [@scharnweber_confessions_2019]. Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al. 2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). Their ability to predict tree performance under drought remains untested.
